---
title: "A Data Analysis of The Storm Data"
author: "Ayako Nagao"
date: "7/7/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## 1.Overview  
In this report, we analyze the NOAA Storm Database to answer the questions about the damages that severe weather events cause across the United States. Namely,

1. Which types of severe weather events are most harmful with respect to population health?  
2. Which types of severe weather events have the greatest economic consequences?  

In the section 2, we load and pre-process the data and in the section 3, we explore the Storm Data in details and answer the question according to our findings.

## 2.Data Processing  

First, we load the data.

```{r loadAndProcess, cache=TRUE,message=FALSE}

url <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'
download.file(url,'stormdata.csv')
data <- read.csv('stormdata.csv')

```

The data has 902297 observations and 37 features.

```{r originalDimension, echo=TRUE}
dim(data)
```

The first 3 rows of data:

```{r headData}
head(data,3)
```

The data has 37 features, but we only need the features that are related to our analysis to answer to the questions.  
After subsetting the features we need, features down to 15.

``` {r subsetFeatures,message=FALSE}

# Remove unnecessary columns
library(dplyr)
data <- select(data,c('STATE__',
                            "STATE","BGN_DATE",'EVTYPE',"LENGTH","WIDTH",
                            "F","MAG","FATALITIES","INJURIES","PROPDMG",
                            "PROPDMGEXP","CROPDMG","CROPDMGEXP","REMARKS"
                            ))

dim(data)
```

And also we only need the observation which caused fatalities/injuries or damages to crops/properties. Thus remove the observations where all of those records equal to zero to make further cleaning process easier.

```{r removeZeroDmg}
# remove data with ALL FATALITIES/INJURIES/CROPDMG/PROPDMG == 0 

noDmgLogical <- (data$FATALITIES == 0)&
  (data$INJURIES == 0)&
  (data$PROPDMG == 0)&
  (data$CROPDMG == 0)

colSums(data[noDmgLogical,c("FATALITIES","INJURIES","PROPDMG","CROPDMG")])

```

It resulted in 254633 rows from 902297.

```{r subsetData}

data <- subset(data, !(noDmgLogical))
dim(data)

```

Next, in order to get the amount of damage to the crops and properties, we need to calculate values in US dollars. The total amounts can be calculated by multiplying PROPDMG/CROPDMG columns by PROPDMGEXP/CROPDMGEXP columns respectively where multipliers are coded in alphabets(B for Billions,M for Millions and so on). Thus, after changing alphabets to multipliers in numbers, make new columns with calculated values.  
I referenced the report [How To Handle Exponent Value of PROPDMGEXP and CROPDMGEXP](https://rstudio-pubs-static.s3.amazonaws.com/58957_37b6723ee52b455990e149edde45e5b6.html) for the clarification.  

``` {r exponentsHandling, message=FALSE}

## Difine exponent column handling function

expFunc <- function(x){
  if ((x=='B') | (x=='b')) {x <- 1e+9
  }else if ((x=='M') | (x=='m')){x <- 1e+6
  }else if((x=='K') | (x=='k')){x <- 1e+3
  }else if((x=='H') | (x=='h')){x <- 1e+2
  }else if(x %in% as.character(0:8)){x <- 1e+1
  }else{x <- 1}
}

# Change alphabets to multiplier
# for CROPDMGEXP
data$CROPDMGEXP <- sapply(data$CROPDMGEXP,expFunc)

# Make a new column for actual damage
data <- mutate(data, totCropDmg=CROPDMG * CROPDMGEXP)

# Repeat the same process for PROPDMGEXP
# for PROPDMGEXP
data$PROPDMGEXP <- sapply(data$PROPDMGEXP, expFunc)
data <- mutate(data, totPropDmg=PROPDMG*PROPDMGEXP)


```


Now, the most important task in this data preprocessing phase is to organize the EVTYPE(event types) column. The official event types are 48. But if we count all unique values in this column, they are 488 at this point. It is due to the mistakes in spelling or being coded with own words. In order to compare all 48 event types, we have to assign incorrect event types to the official ones, meaning reduce 488 categories to 48. The approach I took was,  
  
1. Change all values to upper cases.(It erase the difference between upper cases and lower cases.)
2. Load the official event type categories, (It was copied from the documentation in PDF file.) and change them also in upper cases.  
3. List the event types which do not match with the official event types.  
4. Then deal with them case by cases. For typos, check the pattern and use the grepl function to correct them together, if it was not sure which categories the values are pointing, read REMARKS and decide which events they possibly belong to.  

```{r cleanUp1}

# EVTYPE has 488 unique values
before <- length(unique(data$EVTYPE))

# change all to upper case then 447 EVTYPE

data$EVTYPE <- toupper(data$EVTYPE)
after <- length(unique(data$EVTYPE))

```

Now unique EVTYPE values are reduced to `r after` from `r before`.  
Let's load official event types and compare them to the values in the EVTYPE column.

```{r cleanUp2,results='asis',message=FALSE}

library(kableExtra)
events <- read.table('events.txt',sep = '\n')
events <- toupper(events$V1)
# display official event list
kbl(matrix(events, nrow=16,dimnames =NULL), caption='Official Event Types',
             type='html')  %>% kable_styling(latex_options = 'hold_position',font_size=10)

```

Make a list of events which are not included in above list.

```{r cleanUp3}

notInListEvents <- unique(data[!(data$EVTYPE %in% events),'EVTYPE'])
notInListEventsCounts <- length(notInListEvents)
head(notInListEvents)
```
In total, `r notInListEventsCounts` events are unlisted.  
Below code will assign the unlisted events to the official event lists.

```{r cleanUp4,cache=TRUE}

data$EVTYPE[grepl("^\\s*TSTM WIND",data$EVTYPE)] <- "THUNDERSTORM WIND"
data$EVTYPE[grepl("^HURRICANE",data$EVTYPE)] <- "HURRICANE (TYPHOON)"  
data$EVTYPE[grepl("^THUNDERSTORM",data$EVTYPE)] <- "THUNDERSTORM WIND"
data$EVTYPE[grepl("FLASH FLOOD",data$EVTYPE)] <- "FLASH FLOOD"
data$EVTYPE[grepl("TORNDAO",data$EVTYPE)|
              grepl('TORNADO',data$EVTYPE)] <- "TORNADO"
data$EVTYPE[grepl("^HIGH WIND",data$EVTYPE)] <- "HIGH WIND" 
data$EVTYPE[grepl("^WIND",data$EVTYPE)|
              grepl("^HIGH$",data$EVTYPE)] <- "HIGH WIND"
data$EVTYPE[grepl("^HEAVY RAIN",data$EVTYPE)] <- "HEAVY RAIN"
data$EVTYPE[grepl("LIGHTNING",data$EVTYPE)] <- "LIGHTNING"
data$EVTYPE[grepl("COLD",data$EVTYPE)&
                     !grepl("EXTREME",data$EVTYPE)&
                     !grepl('SNOW',data$EVTYPE)] <- "COLD/WIND CHILL"
data$EVTYPE[grepl("^FLOOD",data$EVTYPE)&
         !grepl("FLASH",data$EVTYPE)] <- "FLOOD"
data$EVTYPE[grepl("EXTREME COLD",data$EVTYPE)|
              grepl("HYPOTHERMIA" ,data$EVTYPE)|
              grepl("LOW TEMPERATURE",data$EVTYPE)] <- "EXTREME COLD/WIND CHILL"
data$EVTYPE[grepl("FREEZE",data$EVTYPE)] <- "FROST/FREEZE"

# Flood has different categories so do Not include other flood categories
data$EVTYPE[grepl("FLOOD",data$EVTYPE)&
                   !grepl("FLASH",data$EVTYPE)&
                     !grepl("COSTAL",data$EVTYPE)&
                     !grepl("LAKESHORE",data$EVTYPE)&
                     !grepl("COASTAL",data$EVTYPE)&
                     !grepl("LAKE",data$EVTYPE)] <-"FLOOD"

data$EVTYPE[grepl("MARINE MISHAP",data$EVTYPE)] <- "MARINE HIGH WIND"
data$EVTYPE[(grepl("^HIGH",data$EVTYPE)& !grepl('WIND',data$EVTYPE)&!grepl("^HIGH$",data$EVTYPE))|
              grepl("HEAVY SEAS",data$EVTYPE)] <- "HIGH SURF"

data$EVTYPE[grepl("SEVERE TURBULENCE",data$EVTYPE)] <- "THUNDERSTORM WIND"
data$EVTYPE[grepl("SNOW",data$EVTYPE)&
                     !(grepl('LAKE',data$EVTYPE))] <- "HEAVY SNOW" 
data$EVTYPE[grepl("APACHE COUNTY",data$EVTYPE)] <- "THUNDERSTORM WIND"
data$EVTYPE[(grepl("TUNDERSTORM",data$EVTYPE)|
                  grepl("^TH.+M",data$EVTYPE)|
                  grepl("SEVERE TH.+M",data$EVTYPE))&
                !grepl("MARINE THUNDERSTORM WIND",data$EVTYPE)] <- "THUNDERSTORM WIND"
data$EVTYPE[grepl("TYPHOON",data$EVTYPE)] <- "HURRICANE (TYPHOON)"
data$EVTYPE[grepl("GUSTY",data$EVTYPE)] <- "STRONG WIND"
data$EVTYPE[grepl("LAKE.+SNOW",data$EVTYPE)] <- "LAKE-EFFECT SNOW"
data$EVTYPE[grepl("HAIL",data$EVTYPE)&
                     !grepl('MARINE',data$EVTYPE)] <- "HAIL" 
data$EVTYPE[grepl("SLIDE",data$EVTYPE)] <- "DEBRIS FLOW"
data$EVTYPE[grepl("TROPICAL",data$EVTYPE)&
                     !grepl('DEPRESSION',data$EVTYPE)] <- "TROPICAL STORM"
data$EVTYPE[grepl("RAIN",data$EVTYPE)] <- "HEAVY RAIN"
data$EVTYPE[grepl("FOG",data$EVTYPE)&
              !grepl('FREEZING',data$EVTYPE)] <- "DENSE FOG"
data$EVTYPE[grepl("COASTAL FLOODING",data$EVTYPE)|
                     grepl('COASTAL FLOODING/EROSION',data$EVTYPE)] <- "COASTAL FLOOD"
data$EVTYPE[grepl("HIGH.+WINDS$",data$EVTYPE)] <- "HIGH WIND"
data$EVTYPE[grepl("FIRE",data$EVTYPE)] <- "WILDFIRE"
data$EVTYPE[grepl("ICE",data$EVTYPE)&
                     !grepl("STORM",data$EVTYPE)] <- "FROST/FREEZE"
data$EVTYPE[grepl("FROST",data$EVTYPE)|
                      grepl("GLAZE",data$EVTYPE)|
                      grepl("ICY ROADS",data$EVTYPE)|
                      grepl("EARLY FROST",data$EVTYPE)] <- "FROST/FREEZE"
data$EVTYPE[grepl("DROUGHT",data$EVTYPE)] <- "DROUGHT"
data$EVTYPE[(grepl("HEAT",data$EVTYPE)&
                     grepl('EX',data$EVTYPE))|
              grepl("HYPERTHERMIA/EXPOSURE",data$EVTYPE)] <- "EXCESSIVE HEAT"
data$EVTYPE[grepl("CHILL",data$EVTYPE)&
                     grepl('EX',data$EVTYPE)] <- "EXTREME COLD/WIND CHILL"
data$EVTYPE[grepl("FLOOD",data$EVTYPE)&
                     grepl('FLASH',data$EVTYPE)] <- "FLASH FLOOD"
data$EVTYPE[grepl("SURF",data$EVTYPE)|
              grepl("ROGUE WAVE",data$EVTYPE)] <- "HIGH SURF"
data$EVTYPE[grepl("WINTER",data$EVTYPE)&
                     !grepl('STORM',data$EVTYPE)] <- "WINTER WEATHER"
data$EVTYPE[grepl("BLIZZARD",data$EVTYPE)] <- "BLIZZARD"
data$EVTYPE[(grepl("WIN[^DG].+",data$EVTYPE)|
                      grepl('COOL',data$EVTYPE)|
                      grepl('FREEZING\\s[^F]',data$EVTYPE))&
                     !grepl('STORM',data$EVTYPE)] <- "WINTER WEATHER"
data$EVTYPE[grepl("HEAVY\\s[PMS][^NUEW]'",data$EVTYPE)|
                     grepl("EXCESSIVE WETNESS",data$EVTYPE)|
                     grepl("PRECIP",data$EVTYPE)] <- "HEAVY RAIN"
data$EVTYPE[grepl("^[^MHST].+WIN[DS]$",data$EVTYPE)|
                     grepl('FORCE',data$EVTYPE)|
              grepl("DRY MIRCOBURST WINDS",data$EVTYPE)] <- "STRONG WIND"
data$EVTYPE[grepl("URBAN",data$EVTYPE)] <- "FLOOD"
data$EVTYPE[grepl("MICROBURST",data$EVTYPE)] <- "STRONG WIND"
data$EVTYPE[grepl("COASTAL.+[N]$",data$EVTYPE)] <- "COASTAL FLOOD"
data$EVTYPE[grepl("HEAT WAVE",data$EVTYPE)|
                     grepl("RECORD HEAT",data$EVTYPE)] <- "EXCESSIVE HEAT"
data$EVTYPE[grepl("WARM",data$EVTYPE)] <- "HEAT"
data$EVTYPE[grepl("WINTER STORMS",data$EVTYPE)] <- "WINTER STORM" 
data$EVTYPE[grepl("STRONG WINDS",data$EVTYPE)] <- "STRONG WIND"
data$EVTYPE[grepl("RIP CURRENTS",data$EVTYPE)] <- "RIP CURRENT"
data$EVTYPE[grepl("WATERSPOUT-",data$EVTYPE)] <- "WATERSPOUT"
data$EVTYPE[grepl("TSTMW",data$EVTYPE)] <- "THUNDERSTORM WIND"
data$EVTYPE[grepl("STORM SURGE",data$EVTYPE)|
              grepl("COASTAL SURGE",data$EVTYPE)] <- "STORM SURGE/TIDE"
data$EVTYPE[grepl("LIG.+G",data$EVTYPE)] <- "LIGHTNING"
data$EVTYPE[grepl("HEAVY SWELLS",data$EVTYPE)|
                     grepl("ROUGH SEAS",data$EVTYPE)] <- "HIGH SURF"
data$EVTYPE[grepl("NON-SEVERE WIND DAMAGE",data$EVTYPE)|
                     grepl("BLOWING DUST",data$EVTYPE)] <- "STRONG WIND" 
data$EVTYPE[grepl("MARINE TSTM WIND",data$EVTYPE)] <- "MARINE THUNDERSTORM WIND"
data$EVTYPE[grepl("LANDSPOUT",data$EVTYPE)] <- "TORNADO"
data$EVTYPE[grepl("DOWNBURST",data$EVTYPE)] <- "STRONG WIND"
data$EVTYPE[grepl("COASTAL STORM",data$EVTYPE)] <- "MARINE THUNDERSTORM WIND"
data$EVTYPE[grepl("ASTRONOMICAL HIGH TIDE",data$EVTYPE)] <- "ASTRONOMICAL LOW TIDE"
data$EVTYPE[grepl("BEACH EROSION",data$EVTYPE)] <- "DEBRIS FLOW"
data$EVTYPE[grepl("RAPIDLY RISING WATER",data$EVTYPE)] <- "FLOOD"
data$EVTYPE[grepl("DROWNING",data$EVTYPE)] <- "HEAVY RAIN"
data$EVTYPE[grepl("AVALANCE",data$EVTYPE)] <- "AVALANCHE"
data$EVTYPE[grepl("HEAVY MIX",data$EVTYPE)] <- "HEAVY SNOW"
data$EVTYPE[grepl("COASTALSTORM",data$EVTYPE)|
              grepl("HEAVY SHOWER",data$EVTYPE)] <- "HEAVY RAIN"
data$EVTYPE[grepl("SLEET/ICE STORM",data$EVTYPE)] <- "SLEET"
data$EVTYPE[grepl("LAKE FLOOD",data$EVTYPE)] <- "LAKESHORE FLOOD"
data$EVTYPE[grepl("LANDSLUMP",data$EVTYPE)] <- "DEBRIS FLOW"
data$EVTYPE[grepl("MARINE ACCIDENT",data$EVTYPE)] <- "HIGH SURF"
data$EVTYPE[grepl("DAM BREAK",data$EVTYPE)] <- "HEAVY RAIN"
data$EVTYPE[grepl("GUSTNADO",data$EVTYPE)] <- "THUNDERSTORM WIND"
data$EVTYPE[grepl("DUST DEVIL WATERSPOUT",data$EVTYPE)] <- "DUST DEVIL"

# remove '?' EVTYPE since no information for weather
data <- subset(data, EVTYPE!='?')

# for "OTHER" type, need to look up REMARKS column 
# get the row index and read all the REMARKS
rownames(data[data$EVTYPE=='OTHER',])

# there were only 3 possible event types. Assign them using row index.
data["249470",'EVTYPE'] <- "AVALANCHE"
data[c("296122","399112","435359"),'EVTYPE'] <- "DUST DEVIL"
data[c("298603","298604","298605","298606","298607","298608","298609",
       "298610", "298611", "298612", "298613", "298614", "298615",
       "298616", "298617", "298618","298619", "298620", "298621",
       "298622" ,"298623", "298624", "298625", "298626", "298627",
       "298628", "298629", "298630", "298631", "298632"),'EVTYPE'] <- "HEAVY RAIN"

# Unlisted events down to 0....!!!!
unique(data$EVTYPE[!(data$EVTYPE %in% events)])


```

Storm Data has the records from 1950 to 2011. There are 48 official categories
of events but from 1950 to 1992, only 1 - 3 events were recorded each year. 
(See appendix 1).Thus we subset the data from 1993.  

```{r subset1993onward, message=FALSE}

# change BGN_DATE column to Date class
library(lubridate)
date <- gsub('/','-',data$BGN_DATE)
date <- gsub(' 0:00:00','',date)

data <- data %>%
  mutate(BGN_DATE=mdy(date),YEAR=year(BGN_DATE))


grouped <- data %>%
  mutate(EVTYPE=factor(EVTYPE),YEAR=factor(YEAR)) %>%
  group_by(YEAR) 


before <- dim(data)[1]
data <- filter(data, YEAR >= '1993')
after <- dim(data)[1]

```

Now data has only records from 1993 to 2011. The number of records reduced to `r after` from `r before`.  



## 3.Results
Table.@ is a list of top 5 weather events which caused the highest fatalities and injuries on US citizens from 1993 to 2011. (see full list for appendix 2).  

```{r result1,message=FALSE}

healthDmgSum <- data %>% 
  group_by(EVTYPE) %>%
  summarize(Fatalities=sum(FATALITIES),Injuries=sum(INJURIES)) %>%
  mutate(Total=Fatalities+Injuries) %>%
  arrange(desc(Total)) 
Rank <- 1:5
kbl(cbind(Rank,healthDmgSum[1:5,]), caption='Top 5 events Caused Highest  Fatalities/Injuries (Descending order of Total)',
             type='html')  %>% 
  kable_styling(latex_options = 'hold_position',full_width=F,position='left') %>%
  column_spec(2,bold=T)%>%
  column_spec(5,bold=T)

```

We grouped the data by year and events, and plotted the total number of fatalities and injuries in Figure.1. 

```{r plotHeath, message=FALSE}
library(ggplot2)
top5healthDmg <- healthDmgSum$EVTYPE[1:5]

PlotHealth <- data %>%
  select(c(YEAR,EVTYPE,FATALITIES,INJURIES)) %>%
  filter(EVTYPE %in% top5healthDmg) %>%
  mutate(TOTAL=FATALITIES+INJURIES) %>%
  group_by(YEAR,EVTYPE) %>%
  summarize(YEARLY=sum(TOTAL))
  
g <- ggplot(PlotHealth,aes(x=YEAR,y=YEARLY,color=EVTYPE))
g <- g + geom_point() 
g + geom_line()

```

Let's look into the details for each event.  
  
1.TORNADO  

```{r parcentageTornado}
# Calculate 2011 tornado percentage in total fatalities/injuries
data2011 <- data %>%
  filter(YEAR=='2011') %>%
  mutate(total=FATALITIES + INJURIES) %>%
  group_by(EVTYPE) %>%
  summarize(S=sum(total))

pct <- round(data2011$S[data2011$EVTYPE=='TORNADO'] / sum(data2011$S),2) * 100

```
 
TORNADO has the highest figure of fatalities and injuries in the records. From 1993, mostly keep the highest place, and 2011 was the one of the worst year, `r pct`% of the total fatalities/injuries in this year were caused by TORNADO.  

2.EXCESSIVE HEAT  
```{r excessiveHeat}
heatFatality <- data %>%
  filter(EVTYPE=='EXCESSIVE HEAT') %>%
  mutate(Total=FATALITIES+INJURIES) %>%
  group_by(STATE) %>%
  summarize(Total=sum(Total)) %>%
  arrange(desc(Total))
kable(heatFatality,type='html')

```

The EXCESSIVE HEAT has the highest fatality among all the events.

3.FLOOD  

In 1998, Oct 17-18 there was a flood event([Central Texas floods](https://en.wikipedia.org/wiki/October_1998_Central_Texas_floods)). The number of the fatalities/injuries in this year occupies nearly 84% of the those of flood events for 20 years, which means this high figure is due to one single event.
4. 
5.LIGHTNING  


```{r economyResult}
economicDmgSum <- data %>%
  group_by(EVTYPE) %>%
  summarize(cropSum=sum(totCropDmg),propSum=sum(totPropDmg)) %>%
  mutate(total=cropSum+propSum) %>%
  arrange(desc(total))



```
## Appendix

1.The number of the unique events by years

```{r apdx1}

# This dataframe shows the number of the unique events each year
countByYear <- summarize(grouped, 'NUMBER OF RECORDED EVENTS'=length(unique(EVTYPE)))
kable(countByYear,type='html')
```

2.The Full list of total Fatalities/Injuries for each event.  

```{r apdx2}

kable(healthDmgSum,type='html')

```

3.The full list of total damages to the crops/properties for each event.
```{r apdx3}

kable(economicDmgSum,type='html')

```

## References
['United States Census Bureau']('https://www.census.gov/library/reference/code-lists/ansi.html')
